Northern Pikeminnow Predation on Juvenile Chinook Salmon in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Populations

### 2021-12-28 analysis/paper/paper.Rmd
start Discussion edits

@@

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately 1.5 kilometer long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is located in a river reach containing historically important overwinter rearing and spring emigration habitat for juvenile Chinook Salmon *O. tshawytscha* and is inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*. Further, the slough is part of the migratory pathway of the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead.

@@

The Deadwater Slough is an approximately 1.5 km section of the mainstem Salmon River located approximately 6 river km downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event that deposited substantial amounts of sediment at the confluence of the Salmon River [@Emerson1973], thereby creating a slow, deep section in the river, spanning approximately 30 acres, and averaging 68 m width.

@@

<!--
Finally, we calculated the proportional stock density (PSD) for Northern Pikeminnow captured during all field efforts, which is a measure of species size structure. PSD is a ratio, typically expressed as a percentage, between the number of "quality-sized" individuals or larger individuals and "stock-sized" individuals. Stock and quality size definitions vary by species. PSD is expressed as:

where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$. We calculated the PSD for Northern Pikeminnow in our study using `r prettyNum(s_size, digits = 0)` mm TL for stock-size and `r prettyNum(q_size, digits = 0)` mm TL for quality-size [@Winther2020].
-->

@@

We estimated an individual, average-size (`r avg_fall_NPM` mm) Northern Pikeminnow to consume `r median(fall_g_summ$cumm_g)` g (range `r min(fall_g_summ$cumm_g)` - `r max(fall_g_summ$cumm_g)` g) of fish during the DSR emigration, depending on the proportion of diet consisting of fish versus invertebrates (Figure \@ref(fig:bio-p)). Similarly, we estimated the average-size (`r avg_spring_NPM` mm) Northern Pikeminnow to consume `r median(spring_g_summ$cumm_g)` g (range `r min(spring_g_summ$cumm_g)` - `r max(fall_g_summ$cumm_g)` g) of fish during the NRR emigration (Figure \@ref(fig:bio-p)). Given an average weight of `r avg_DSR_g` g for DSR emigrants, that would equate to `r round(median(adult_impact$juv_chnk[adult_impact$season == "Fall"]), 2)` (range `r round(min(adult_impact$juv_chnk[adult_impact$season == "Fall"]), 2)` - `r round(max(adult_impact$juv_chnk[adult_impact$season == "Fall"]), 2)`) fish consumed by an individual during fall and an additional `r round(median(adult_impact$juv_chnk[adult_impact$season == "Spring"]), 2)` (range `r round(min(adult_impact$juv_chnk[adult_impact$season == "Spring"]), 2)` - `r round(max(adult_impact$juv_chnk[adult_impact$season == "Spring"]), 2)`) fish consumed using an average weight of `r avg_NRR_g` g for NRR emigrants.

Given a mean population size of `r prettyNum(unique(adult_impact$N[adult_impact$season == "Fall"]), big.mark = ",")` from fall estimates and an estimate of `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Spring"]), 0), big.mark = ",")` for spring 2021, and assuming the median value from diet scenarios (i.e, 50% of their diet consists of fish), we'd estimate that Northern Pikeminnow can consume `r prettyNum(round(median(adult_impact$smolt_N[adult_impact$season == "Fall"]), 0), big.mark = ",")` (95% CI: `r prettyNum(round(min(adult_impact$smolt_N[adult_impact$season == "Fall"]), 0), big.mark = ",")` - `r prettyNum(round(max(adult_impact$smolt_N[adult_impact$season == "Fall"]), 0), big.mark = ",")`) fish during the DSR emigration and `r prettyNum(round(median(adult_impact$smolt_N[adult_impact$season == "Spring"]), 0), big.mark = ",")` (95% CI: `r prettyNum(round(min(adult_impact$smolt_N[adult_impact$season == "Spring"]), 0), big.mark = ",")` - `r prettyNum(round(max(adult_impact$smolt_N[adult_impact$season == "Spring"]), 0), big.mark = ",")`) fish during the NRR emigration in Deadwater Slough. Table \@ref(tab:adult-imp) shows the estimated number of fish consumed by an individual Northern Pikeminnow and by the population across all diet scenarios during both the fall (DSR) and spring (NRR) peak emigration seasons.

@@

We estimated the population size of Northern Pikeminnow in the Deadwater Slough to be greater than ***xx,xxx*** during the fall emigration period for DSR Chinook Salmon. That estimate translates to a density of ***xxx*** Northern Pikeminnow per 100 m or ***xxx*** per 100 m^2^ which is ***similar/more/less*** than estimates from elsewhere in the Columbia River ***(citation)*** where substantial Northern Pikeminnow predation impacts on salmonids have led to bounty programs aimed at reducing Northern Pikeminnow abundance. The population size of Northern Pikeminnow was not directly estimated during the spring NRR Chinook salmon emigration period; however, the relative abundance measured at CPUE was comparable to the fall sampling periods ***(update statement later)***. The population of Northern Pikeminnow in Deadwater Slough was estimated to consume between ***xx,xxx*** and ***xx,xxx*** juvenile Chinook Salmon during the ***x*** sampling periods and result in an estimated reduction of returning adults between ***xxx*** and ***x,xxx***. We suggest that the habitat modifications that created the Deadwater Slough have resulted in favorable conditions for Northern Pikeminnow, including improved conditions for predation upon juvenile Chinook Salmon ***(add detail here)***. Therefore, predation by Northern Pikeminnow in the Deadwater Slough likely has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon River MPG.

<!--
The PSD for Northern Pikeminnow in our study and across all three years was `r prettyNum(psd, digits = 1)`% i.e., `r prettyNum(psd, digits = 1)` of captured fish were over the quality size classification of `r prettyNum(q_size, digits = 0)` mm TL defined by @Winther2020. For comparison, @Winther2020 reported a PSD for Northern Pikeminnow captured below Bonneville Dam of 41% and 18% in Bonneville Reservoir, suggesting that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to population reported elsewhere in the Columbia River. 
-->

@@

@@

## Conclusions

*	We have also presented a novel modeling framework for estimating predation on native, critically endangered anadromous species which can be applied to other areas of interest. John day? Others?
*	The end.

@@

```{r adult-imp}
adult_impact %>%
  select(season,
         perc_diet,
         juv_chnk,
         smolt_N,
         smolt_Lci,
         smolt_Uci,
         medsar,
         sdsar,
         adult_N,
         adult_Lci,
         adult_Uci) %>%
  mutate(juv_chnk = round(juv_chnk, 2),
         smolt_N = round(smolt_N, 0),
         smolt_Lci = round(smolt_Lci, 0),
         smolt_Uci = round(smolt_Uci, 0),
         medsar = round(medsar, 5),
         sdsar = round(sdsar, 5),
         adult_N = round(adult_N, 0),
         adult_Lci = round(adult_Lci, 0),
         adult_Uci = round(adult_Uci, 0)) %>%
  rename(Effort = season,
         `p Fish` = perc_diet,
         `Juv Chinook` = juv_chnk,
         `Juv N` = smolt_N,
         `Juv L CI` = smolt_Lci,
         `Juv U CI` = smolt_Uci,
         `Median SAR` = medsar,
         `SD SAR` = sdsar,
         `Adult N` = adult_N,
         `Adult L CI` = adult_Lci,
         `Adult U CI` = adult_Uci) %>%
  kable(booktabs = T,
        format.args = list(big.mark = ","),
        caption = "Estimates of the number of DSR (fall) and NRR (spring) emigrants consumed for each season and diet scenario. The diet scenario indicates the proportion of the Northern Pikeminnow diet consisting of fish. The 'Juv Chinook' column indicates the number of juvenile Chinook consumed by an individual Northern Pikeminnow during the fall or spring seasons and is taken from each of the 18 bioenergetics model runs. The number of smolts consumed (and 95% confidence intervals) are calculated using Northern Pikeminnow abundance estimates of 28,144 (95% CI: 14,378 - 59,978) for fall and 14,897 (95% CI: 7,610 - 31,747) for spring. The number of 'adult equivalents' is also estimated using a Lower Granite Dam smolt-to-adult rate (SAR) from McCann et al. 2019.") %>%
  kable_styling(font_size = 8)

>>>>

```{r deadwater-attr}
# set some deadwater attributes
ddw_rkms = 1.871
ddw_m2 = 119431
ddw_avg_width_m = 68
ddw_acres = ddw_m2 / 4047
ddw_hectares = ddw_m2 / 10000

@@

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` kilometer long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is located in a river reach containing historically important overwinter rearing and spring emigration habitat for juvenile Chinook Salmon *O. tshawytscha* and is inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*. Further, the slough is part of the migratory pathway of the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead.

@@

The Deadwater Slough is an approximately 1.5 km section of the mainstem Salmon River located approximately 6 river km downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event that deposited substantial amounts of sediment at the confluence of the Salmon River [@Emerson1973], thereby creating a slow, deep section in the river, spanning approximately `r round(ddw_acres)` acres (`r round(ddw_hectares)` hectares), and averaging `r round(ddw_avg_width_m)` m width.

@@

@@

We estimated an individual, average-size (`r avg_fall_NPM` mm) Northern Pikeminnow to consume `r median(fall_g_summ$cumm_g)` g (range `r min(fall_g_summ$cumm_g)` - `r max(fall_g_summ$cumm_g)` g) of fish during the DSR emigration to maintain their body size, depending on the proportion of diet consisting of fish versus invertebrates (Figure \@ref(fig:bio-p)). Similarly, we estimated an average-size (`r avg_spring_NPM` mm) Northern Pikeminnow to consume `r median(spring_g_summ$cumm_g)` g (range `r min(spring_g_summ$cumm_g)` - `r max(fall_g_summ$cumm_g)` g) of fish during the NRR emigration (Figure \@ref(fig:bio-p)). Given an average weight of `r avg_DSR_g` g for DSR emigrants, that would equate to `r round(median(adult_impact$juv_chnk[adult_impact$season == "Fall"]), 2)` (range `r round(min(adult_impact$juv_chnk[adult_impact$season == "Fall"]), 2)` - `r round(max(adult_impact$juv_chnk[adult_impact$season == "Fall"]), 2)`) fish consumed by an individual during fall and an additional `r round(median(adult_impact$juv_chnk[adult_impact$season == "Spring"]), 2)` (range `r round(min(adult_impact$juv_chnk[adult_impact$season == "Spring"]), 2)` - `r round(max(adult_impact$juv_chnk[adult_impact$season == "Spring"]), 2)`) fish consumed using an average weight of `r avg_NRR_g` g for NRR emigrants.

Given a mean Northern Pikeminnow population size of `r prettyNum(unique(adult_impact$N[adult_impact$season == "Fall"]), big.mark = ",")` from fall estimates and an estimate of `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Spring"]), 0), big.mark = ",")` for spring 2021 and using the median value from diet scenarios (i.e, 50% of their diet consists of fish), we'd estimate that Northern Pikeminnow can consume `r prettyNum(round(median(adult_impact$smolt_N[adult_impact$season == "Fall"]), 0), big.mark = ",")` (95% CI: `r prettyNum(round(min(adult_impact$smolt_N[adult_impact$season == "Fall"]), 0), big.mark = ",")` - `r prettyNum(round(max(adult_impact$smolt_N[adult_impact$season == "Fall"]), 0), big.mark = ",")`) fish during the DSR emigration and `r prettyNum(round(median(adult_impact$smolt_N[adult_impact$season == "Spring"]), 0), big.mark = ",")` (95% CI: `r prettyNum(round(min(adult_impact$smolt_N[adult_impact$season == "Spring"]), 0), big.mark = ",")` - `r prettyNum(round(max(adult_impact$smolt_N[adult_impact$season == "Spring"]), 0), big.mark = ",")`) fish during the NRR emigration in Deadwater Slough. Table \@ref(tab:adult-imp) shows the estimated number of fish consumed by an individual Northern Pikeminnow and by the population across all diet scenarios during both the fall (DSR) and spring (NRR) peak emigration seasons.

@@

```{r john-day}
john_day_res_abund = 85316
john_day_rkms = 123
john_day_dens = john_day_res_abund / john_day_rkms

low_columbia_fish_km = 2814

The mean population size of Northern Pikeminnow in the Deadwater Slough during the fall season was estimated to be `r prettyNum(unique(adult_impact$N[adult_impact$season == "Fall"]), big.mark = ",")` and during the spring 2021 to be `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Spring"]), 0), big.mark = ",")`. Those estimates translate to linear densities of `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Fall"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Spring"]) / ddw_rkms), big.mark = ",")` which is `r round(unique(adult_impact$N[adult_impact$season == "Fall"]) / ddw_rkms / john_day_dens)` and `r round(unique(adult_impact$N[adult_impact$season == "Spring"]) / ddw_rkms / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r as.character(english(round(unique(adult_impact$N[adult_impact$season == "Fall"]) / ddw_rkms / low_columbia_fish_km)))` and `r as.character(english(round(unique(adult_impact$N[adult_impact$season == "Spring"]) / ddw_rkms / low_columbia_fish_km)))` times greater than projections for the lower Columbia River [@Beamesderfer1996]. These densities for Deadwater Slough initially seem unrealistic especially given the size (volume) of the Salmon River at the slough relative to the Columbia River. However, we did catch `r prettyNum(nrow(cmr_df), big.mark = ",")` Northern Pikeminnow over the course of the study, of which only `r sum(cmr_df$capture_type == "Recapture", na.rm = T)` fish (one in every `r round(nrow(cmr_df) / sum(cmr_df$capture_type == "Recapture", na.rm = T))`) were recaptures. Further, @Beamesderfer1991 did acknowledge that they were unsuccessful in finding a suitable method for sampling offshore; they noted that water velocity, depth, irregular bottom contours, and barge traffic made sampling offshore ineffective and so their abundance estimates only apply to nearshore areas. This was not the case at Deadwater Slough where maximum depth was approximately 10-15 m with a relatively homogeneous bottom contour i.e., angling was effective for the entire area. Our estimates of Northern Pikeminnow areal density are also higher than those reported elsewhere. We estimate the density of Northern Pikeminnow in Deadwater Slough, as fish per hectare, to be `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Fall"]) / ddw_hectares), big.mark = ",")` for fall and `r prettyNum(round(unique(adult_impact$N[adult_impact$season == "Spring"]) / ddw_hectares), big.mark = ",")` for spring. For comparison, the only other available estimates of areal density, that the authors are aware of, are 4.4 fish per hectare for John Day Reservoir [@Beamesderfer1991] and 15 fish per hectare in Lake Washington [@Bartoo1977].

Northern Pikeminnow in Deadwater Slough were not only abundant, but anglers noticed that many were also large-bodied. For that reason, we calculated the proportional stock density (PSD) for Northern Pikeminnow in Deadwater Slough, which is a measure of species size structure. PSD is a ratio, typically expressed as a percentage, between the number of "quality_sized" (larger) individuals and "stock-sized" individuals.

where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$. Stock and quality size definitions vary by species. We calculated the PSD for Northern Pikeminnow in Deadwater Slough using `r prettyNum(s_size, digits = 0)` mm TL for stock-size and `r prettyNum(q_size, digits = 0)` mm TL for quality-size [@Winther2020]. The PSD for Northern Pikeminnow in Deadwater Slough across all three efforts was `r prettyNum(psd, digits = 1)`% i.e., `r prettyNum(psd, digits = 1)`% of captured fish were over the quality size classification defined by @Winther2020. For comparison, @Winther2020 reported a PSD for Northern Pikeminnow captured below Bonneville Dam of 41% and 18% in Bonneville Reservoir, suggesting that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to populations reported elsewhere in the Columbia River. In fact, the state record Northern Pikeminnow, which measured `r max(cmr_df$length, na.rm = T)` mm (`r round(max(cmr_df$length, na.rm = T) / 25.4, 1)` in) TL, was caught in Deadwater Slough during the fall 2020 effort.

@@

The population of Northern Pikeminnow in Deadwater Slough was estimated to consume between ***xx,xxx*** and ***xx,xxx*** juvenile Chinook Salmon during the ***x*** sampling periods and result in an estimated reduction of returning adults between ***xxx*** and ***x,xxx***.

@@

projections for the lower Columbia River (Beamesderfer et al. 1996) where substantial Northern Pikeminnow predation impacts on salmonids have led to bounty programs aimed at reducing Northern Pikeminnow abundance.

## Conclusions
* We suggest that the habitat modifications that created the Deadwater Slough have resulted in favorable conditions for Northern Pikeminnow, including improved conditions for predation upon juvenile Chinook Salmon ***(add detail here)***. Therefore, predation by Northern Pikeminnow in the Deadwater Slough likely has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon River MPG.
*	We have also presented a novel modeling framework for estimating predation on native, critically endangered anadromous species which can be applied to other areas of interest. John day? Others?
*	The end.

@@

```{r adult-imp}
adult_impact %>%
  select(season,
         perc_diet,
         juv_chnk,
         smolt_N,
         smolt_Lci,
         smolt_Uci,
         medsar,
         sdsar,
         adult_N,
         adult_Lci,
         adult_Uci) %>%
  mutate(juv_chnk = round(juv_chnk, 2),
         smolt_N = round(smolt_N, 0),
         smolt_Lci = round(smolt_Lci, 0),
         smolt_Uci = round(smolt_Uci, 0),
         medsar = round(medsar, 5),
         sdsar = round(sdsar, 5),
         adult_N = round(adult_N, 0),
         adult_Lci = round(adult_Lci, 0),
         adult_Uci = round(adult_Uci, 0)) %>%
  rename(Season = season,
         `p Fish` = perc_diet,
         `Fish Per NPM` = juv_chnk,
         `Fish N` = smolt_N,
         `Fish L CI` = smolt_Lci,
         `Fish U CI` = smolt_Uci,
         `Median SAR` = medsar,
         `SD SAR` = sdsar,
         `Adult N` = adult_N,
         `Adult L CI` = adult_Lci,
         `Adult U CI` = adult_Uci) %>%
  kable(booktabs = T,
        format.args = list(big.mark = ","),
        caption = "Estimates of the number of fish consumed for each season and diet scenario. The diet scenario indicates the proportion of the Northern Pikeminnow diet consisting of fish. The 'Fish Per NPM' column indicates the number of fish an individual Northern Pikeminnow to consume during the fall or spring seasons assuming an average weight of 10.3 (DSR) and 10.9 (NRR) grams for juvenile Chinook Salmon and is taken from each of the 18 bioenergetics model runs. The number of fish consumed (and 95% confidence intervals) are calculated using Northern Pikeminnow abundance estimates of 28,144 (95% CI: 14,378 - 59,978) for fall and 14,897 (95% CI: 7,610 - 31,747) for spring. The number of 'adult equivalents' is also estimated using a Lower Granite Dam smolt-to-adult rate (SAR) for Chinook Salmon from McCann et al. 2019.") %>%
  kable_styling(font_size = 8)
